Regression Cube Analysis of Cohort Study Data

### 2015-03-18 draft.tex
first draft of whole structure

%% Abstract section.
\abstract{%
\com{Problem}.
Epidemiological studies comprise of heterogenous data about a subject group (a \emph{cohort}) to define disease-specific risk factors.
%%
These data contain information (\emph{features}) about a subjects lifestyle, medical conditions and also medical images of the whole body.
%%
These features are analyzed using statistical regression analyses towards features indicating a disease (the \emph{target feature}).
%%
%These complex data are hard to analyze concurrently, only a few features can be considered in statistical analyses, such as regression analyses.
These analyses usually include two to five features (\emph{independent features}) towards the disease of interest.
\com{New Solution}.
%(New Solution).
We propose a new analyses approach of epidemiological data sets by incorporating all dimensions in a exhaustive regression-based analyses.
%%
It takes all possible combinations of \emph{independent features} towards a \emph{target feature} into account and provides a visualization, which allows for insights into the data by showing relationships possibly unknown to the domain expert.
%%
%The most meaningful features 
A 3D-visualization of all possible combinations of two to three target variables acts as overview over the whole data set.
%%
We call it the \emph{Regression Cube}.
%%
Slicing through the regression cube allows for the detailed analysis of features towards the target disease.
%%
The regression formula can be adjusted by the user to describe disease-specific hypotheses.
%%
Influences of features can be assessed using a difference view, comparing different calculation results.
\com{Validation}.
%%
We applied our \emph{Regression Cube} method to a hepatic steatosis data set to reproduce results from a data-mining driven analysis.
%%
We conducted a qualitative analysis with three domain experts on a breast fat data set to derive insights into the correlation between breast lesions and non-image variables.
%%
\com{Results}. TODO.
%%
\com{Implications}.
Epidemiological data sets can for the first time be visually overviewed using a regression-based analysis of all features towards a disease. 
%With our technique it is possible for the first time to provide a overview visualization of variables regarding a target variable.

@@

\subsection{The Study of Health in Pomerania (SHIP)}
After the pioneering Rotterdam study (started in 1990), several MR imaging study initiatives were initiated.
%
They slightly differ in clinical focus, acquired data and epidemiological research questions.
%
Starting in 1997 with a cohort of 4,308 subjects, the \texttt{SHIP}, located in Northern Germany, aims to characterize health and disease in the widest range possible \cite{Volzke2011}.
%
Data are collected without focus on a group of diseases.
%
This allows to query the data regarding many diseases and conditions.
%
Subjects were examined in a 5-year time span, continuously adding new parameters including MRI scans in the last iteration \cite{Hegenscheid2009}.
%
\section{Prior and Related Work}

\paragraph{Visual Analysis of Heterogenous Data.}
Zhang et al. \cite{Zhang2012} provide a web-based system for analyzing subject groups with linked views and batch-processing capabilities for categorizing new subject entries into the data set.
%
Their definition of a cohort differs from the understanding of the term in an epidemiological context by denoting every parameter-divided subject group as individual cohort.
%
Due to the short paper length, detail is missing on the data types and their algorithms of identifying similar subjects or whether they employ statistical measures.
%
We employ the idea of adding variables via drag and drop into a canvas area.

Generalized Pairs Plots (\texttt{GPLOM´S}) are an information visualization technique comparing heterogenous variables pairwise using a plot-matrix grouped by type \cite{GPLOMS, Francois2013}.
%%
They are useful to gain an overview over numerous variables and their distributions.
%%
Histograms, bar charts, scatter plots and heat maps are used to visualize variable combinations with regard to their type.
%%
The resulting matrix provides an \emph{overview visualization}, but requires a lot of screen space for many variables (127 in our application scenario).
%%
We incorporate the idea of adaptive type-dependent visualizations.
%%
Dai et al. \cite{Dai2005} explored risk factors by incorporating choropleth maps of epidemiological variables (e.g., mortality rates in a region) with parallel coordinates, bar charts and scatter plots with integrated regression lines.
%%
Their findings yielded a \emph{Concept Map}, which linked cancer-related associations via graph edges.
%%
While their goal to identify possible risk factors using socio-economic and health data is similar to ours, they focus on iteratively refining defined hypotheses and on geographical data.
%%
We employ the use of small multiples for incorporating heterogenous data types for comparability.
%%
Chui et al. \cite{Chui2011} visualized associations in time-dependent epidemiological data using time-series plots highlighting risk factor differences in age and gender.
%%
%The connection of time series plots with histograms and heat maps allows for fast insight into complex relationships and 
While the work shows how different visualization techniques provide insight into these data sets, it focuses on the time aspect, which is not present for our data.
%%
% *********************** Our own Work ***********************
% Prior Work
\paragraph{Prior Work.}
We visualized lumbar spine variabilities based on a semi-automatic shape detection algorithm of 490 participants of the \texttt{SHIP-2} cohort \cite{Klemm2013VMV}.
%
Hierarchical agglomerative clustering divided the population into shape-related groups.
%
As proof of concept, a relation between the size of the segmented shape and the measured size of the subjects was shown.
%
This work focuses on incorporating these derived data as new variables, enabling to include it into the hypothesis validation and generation process.
%
When applying clustering techniques to the non-image data it was found that \texttt{k-Prototypes} and \texttt{DBSCAN} are appropriate, but are strongly dependent on the chosen variables and distance measures \cite{Klemm2014BVM}.
%
Niemann et al. \cite{Niemann2014} presented an interactive data mining tool for the assessment of risk factors of hepatic steatosis, the fatty liver disease.
%
Association rules created by data mining methods can be analyzed interactively with their tool and highlight potentially overlooked variables.
% *********************** / Our own Work ***********************
\section{Regression Cubes} \label{Image Centric Cohort Study Data in Interactive Visual Analysis Context}
%%
\com{Target Group: Physicians/Epidemiologists with basic statistical background to understand regression formulas and Interaction}
\subsection{Analysis Workflow}
\com{Different Formulas work as different steps in the analysis}
%%
\section{System Design and Implementation} \label{Interaction- and Visualization Techniques}
%
\subsection{Design and Visualization Techniques} \label{Structure and Workflow}

@@

\subsection{Implementation} \label{implementation}
%%
\com{Use VAST'14 as Guideline}
%
\section{Application} \label{application}
%
\subsection{The Breast Fat Data Set}

@@

\subsection{Participants, Setup and Procedure}
%
\paragraph{Setup.} Due to the large geographical distance, the evaluation was done completely web-based.
%%
\paragraph{Procedure.}
%%
\com{Analyses here ...}
%%
\subsection{Further Feedback and Lessons Learned} \label{Lessons Learned}
%%

\section{Summary and Conclusion}
%%
\begin{small}
	\acknowledgments{Omitted due to blind review}
% \acknowledgments{SHIP is part of the Community Medicine Research net of the University of Greifswald, Germany, which is funded by the Federal Ministry of Education and Research (grant no. 03ZIK012), the Ministry of Cultural Affairs as well as the Social Ministry of the Federal State of Mecklenburg-West Pomerania. Whole-body MR imaging was supported by a joint grant from Siemens Healthcare, Erlangen, Germany and the Federal State of Mecklenburg-Vorpommern. The University of Greifswald is a member of the ‘Centre of Knowledge Interchange’ program of the Siemens AG. This work was supported by the DFG Priority Program 1335: Scalable Visual Analytics. We thank Marko Rak and Klaus Toennies for providing the image detection data.}
\end{small}
\clearpage
\newpage
\bibliographystyle{abbrv}
%%use following if all content of bibtex file should be shown
%\nocite{*}
\bibliography{bibliography}
\end{document}

>>>>

%% Abstract section.
\abstract{%
\com{Problem}.
Epidemiological studies comprise heterogeneous data about a subject group (a \emph{cohort}) to define disease-specific risk factors.
%%
These data contain information (\emph{features}) about a subjects' lifestyle, medical conditions and also medical images of the whole body.
%%
These features are analyzed using statistical regression analyses towards features indicating a disease (the \emph{target feature}).
%%
%These complex data are hard to analyze concurrently, only a few features can be considered in statistical analyses, such as regression analyses.
These analyses usually include two to five features (\emph{independent features}) towards the disease of interest.
\com{New Solution}.
%(New Solution).
We propose a new analyses approach of epidemiological data sets by incorporating all dimensions in an exhaustive regression-based analyses.
%%
%It takes all possible combinations of \emph{independent features} towards a \emph{target feature} into account and provides a visualization, which allows for insights into the data by showing relationships possibly unknown to the domain expert.
This approach combines all \emph{independent features} towards a \emph{target feature} and provides a visualization, which allows for insights into the data by showing the domain expert possibly unknown relationships.
%%
%The most meaningful features 
%A 3D-visualization of all possible combinations of two to three target variables acts as overview over the whole data set.
A 3D-visualization of all combinations of two to three target variables acts as an overview over the whole data set.
%%
We call it the \emph{Regression Cube}.
%%
Slicing through the regression cube allows for the detailed analysis of features towards the target disease.
%%
The regression formula can be adjusted by the user to describe disease-specific hypotheses.
%%
Influences of features can be assessed using a difference view, comparing different calculation results.
\com{Validation}.
%%
We applied our \emph{Regression Cube} method to a hepatic steatosis data set to reproduce results from a data-mining driven analysis.
%%
We conducted a qualitative analysis with three domain experts on a breast fat data set to derive insights into the correlation between breast lesions and non-image variables.
%%
\com{Results}. TODO.
%%
\com{Implications}.
Epidemiological data sets can for the first time be visually overviewed using a regression-based analysis of all features towards a disease. 
%With our technique it is possible for the first time to provide a overview visualization of variables regarding a target variable.

@@

\subsection{Regression Analysis}
\com{Regression cubes are an important tool for checking statistical resilience.}
\com{Describe how regression cubes work (including notation).}
\com{Move this to Regression Cube part?}

\subsection{The Study of Health in Pomerania (SHIP)}
After the pioneering Rotterdam study (started in 1990), several MR imaging study initiatives were initiated.
%
They slightly differ in clinical focus, acquired data and epidemiological research questions.
%
Starting in 1997 with a cohort of 4,308 subjects, the \texttt{SHIP}, located in Northern Germany, aims to characterize health and disease in the widest range possible \cite{Volzke2011}.
%
Data are collected without focus on a group of diseases.
%
This allows to query the data regarding many diseases and conditions.
%
Subjects were examined in a 5-year time span, continuously adding new parameters including MRI scans in the last iteration \cite{Hegenscheid2009}.
%
\section{Prior and Related Work}
\com{From VAST'14 Paper:}
\paragraph{Visual Analysis of Heterogenous Data.}
Zhang et al. \cite{Zhang2012} provide a web-based system for analyzing subject groups with linked views and batch-processing capabilities for categorizing new subject entries into the data set.
%
Their definition of a cohort differs from the understanding of the term in an epidemiological context by denoting every parameter-divided subject group as individual cohort.
%
Due to the short paper length, detail is missing on the data types and their algorithms of identifying similar subjects or whether they employ statistical measures.
%
We employ the idea of adding variables via drag and drop into a canvas area.

Generalized Pairs Plots (\texttt{GPLOM´S}) are an information visualization technique comparing heterogenous variables pairwise using a plot-matrix grouped by type \cite{GPLOMS, Francois2013}.
%%
They are useful to gain an overview over numerous variables and their distributions.
%%
Histograms, bar charts, scatter plots and heat maps are used to visualize variable combinations with regard to their type.
%%
The resulting matrix provides an \emph{overview visualization}, but requires a lot of screen space for many variables (127 in our application scenario).
%%
We incorporate the idea of adaptive type-dependent visualizations.
%%
Dai et al. \cite{Dai2005} explored risk factors by incorporating choropleth maps of epidemiological variables (e.g., mortality rates in a region) with parallel coordinates, bar charts and scatter plots with integrated regression lines.
%%
Their findings yielded a \emph{Concept Map}, which linked cancer-related associations via graph edges.
%%
While their goal to identify possible risk factors using socio-economic and health data is similar to ours, they focus on iteratively refining defined hypotheses and on geographical data.
%%
We employ the use of small multiples for incorporating heterogenous data types for comparability.
%%
Chui et al. \cite{Chui2011} visualized associations in time-dependent epidemiological data using time-series plots highlighting risk factor differences in age and gender.
%%
%The connection of time series plots with histograms and heat maps allows for fast insight into complex relationships and 
While the work shows how different visualization techniques provide insight into these data sets, it focuses on the time aspect, which is not present for our data.
%%
% *********************** Our own Work ***********************
% Prior Work
\paragraph{Prior Work.}
We visualized lumbar spine variabilities based on a semi-automatic shape detection algorithm of 490 participants of the \texttt{SHIP-2} cohort \cite{Klemm2013VMV}.
%
Hierarchical agglomerative clustering divided the population into shape-related groups.
%
As proof of concept, a relation between the size of the segmented shape and the measured size of the subjects was shown.
%
This work focuses on incorporating these derived data as new variables, enabling to include it into the hypothesis validation and generation process.
%
When applying clustering techniques to the non-image data it was found that \texttt{k-Prototypes} and \texttt{DBSCAN} are appropriate, but are strongly dependent on the chosen variables and distance measures \cite{Klemm2014BVM}.
%
Niemann et al. \cite{Niemann2014} presented an interactive data mining tool for the assessment of risk factors of hepatic steatosis, the fatty liver disease.
%
Association rules created by data mining methods can be analyzed interactively with their tool and highlight potentially overlooked variables.
% *********************** / Our own Work ***********************
\section{Regression Cube Analysis of Cohort Study Data}
%%
\com{Target Group: Physicians/Epidemiologists with basic statistical background to understand regression formulas and Interaction.}

\subsection{Target-Variable-dependent Dimension Reduction}
\com{Explain the CFS algorithm and how it is applied to reduce the number of dimensions and therefore the calculation process.}

\subsection{Cube Description using Regression Formula Notation}
\com{Regression formulas are described using standard regression formula notation by introducing variables.}

\subsection{Abstracting Regression Results using R$^2$}
\com{R$^2$ are used to encode regression results to color spaces.}
\com{R$^2$ is comparable between regression results of the same type, but not in between types. Therefore it is encoded using different colours.}

\subsection{Analysis Workflow}
\begin{figure}[htb]
 \centering
 \includegraphics[width=2.8in]{figures/placeholder}
 \caption{
 \com{Workflow for different analysis approaches.}
 \com{a) Analysis without hypothesis about the data set (z ~ x+y)}
 \com{b) Analysis without hypothesis with confounder (z ~ age)}
 \com{c) Analysis with hypothesis about the data set (Var ~ x+y+z)}
 \com{d) Clarify using confounder (Var ~ x)}
 }
  \label{fig:Workflow}
\end{figure}
\com{Different Formulas work as different steps in the analysis (Fig.~\ref{fig:Workflow})}
%%
\section{System Design and Implementation} \label{Interaction- and Visualization Techniques}
%%

\subsection{Design and Visualization Techniques} \label{Structure and Workflow}
\begin{figure}[htb]
 \centering
 \includegraphics[width=2.8in]{figures/placeholder}
 \caption{
 \com{Screenshot of the cube visualization embedded in the framework}
 \com{Also include active tooltip}
 }
  \label{fig:Cube}
\end{figure}

\begin{figure}[htb]
 \centering
 \includegraphics[width=2.8in]{figures/placeholder}
 \caption{
 \com{Screenshot of the cube visualization embedded in the framework}
 }
  \label{fig:comparison}
\end{figure}
%%
\com{Simplicity of interface to allow for steep learning curve.}
\com{Visualization as skewed cube to reduce visual clutter and remove double entries.}
\com{Cube acts as visual mini map to give a global impression over the data.}
\com{current pane shown using a heat map visualization.}
\com{details on demand for selected regression formula.}
\com{comparison view using reference cubes.}

@@

\subsection{Implementation} \label{implementation}
\begin{figure}[htb]
 \centering
 \includegraphics[width=2.8in]{figures/placeholder}
 \caption{
 \com{Describe the Bootstrap/Angular/D3 Frontend, Node backend, R/OCPU Backend.}
 }
  \label{fig:Implementation}
\end{figure}
%%
\com{Use VAST'14 as Guideline}
\com{Speed is merely a matter of available server machines due to parallelization process.}
\com{Security enabled by hashing files.}
\com{Bootstrap/Angular/D3 Frontend, Node backend, R/OCPU Backend (Fig.~\ref{fig:Implementation})}

\section{Application} \label{application}
%
\subsection{The Breast Fat Data Set}
\com{Describe how the data was acquired and preprocessed}

@@

\subsection{The Hepatic Steatosis Data Set}
\com{Uli: Describe the data set and also prior work on it.}

\subsection{Participants, Setup and Procedure}
%%
\com{Use of VDAR Technique}
%
\paragraph{Setup.} Due to the large geographical distance, the evaluation was done completely web-based.
%%
\paragraph{Procedure.}
%%
\subsection{Case 1: Hypothesis-free Analysis of the Breast Cancer Data Set}

\subsection{Case 2: Hypothesis-driven Analysis of the Hepatic Steatosis Data Set}
%%
\subsection{Further Feedback and Lessons Learned} \label{Lessons Learned}
%%
\com{Feedback from the evaluation goes here.}
\com{Time-aspect critical, interactive analysis requires for fast response. The method needs to be speeded up.}

\section{Summary and Conclusion}
\paragraph{Future Work.}
\com{Implementing of regression analysis on the graphics card.}
%%
\begin{small}
	\acknowledgments{Omitted due to blind review}
% \acknowledgments{SHIP is part of the Community Medicine Research net of the University of Greifswald, Germany, which is funded by the Federal Ministry of Education and Research (grant no. 03ZIK012), the Ministry of Cultural Affairs as well as the Social Ministry of the Federal State of Mecklenburg-West Pomerania. Whole-body MR imaging was supported by a joint grant from Siemens Healthcare, Erlangen, Germany and the Federal State of Mecklenburg-Vorpommern. The University of Greifswald is a member of the ‘Centre of Knowledge Interchange’ program of the Siemens AG. This work was supported by the DFG Priority Program 1335: Scalable Visual Analytics. We thank Marko Rak and Klaus Toennies for providing the image detection data.}
\end{small}
\clearpage
\newpage
\bibliographystyle{abbrv}
%%use following if all content of bibtex file should be shown
%\nocite{*}
\bibliography{bibliography}
\end{document}
